Domestic cooking and likelihood

### 2021-02-01 inst/cook.Rmd
fixes #75

```{r loadlib}
library(hyper2)
H <- hyper2(pnames=paste("p",1:5,sep=""))
H
```

@@

```{r data}
H <- H +
    rank_likelihood(c(5,2,3,4)) + # person 1 says 5 is the best, 4 the worst.
    rank_likelihood(c(1,5,3,4)) + # person 2 says 1 is the best, 4 the worst.
    rank_likelihood(c(1,4,2,5)) + # person 3 says 1 is the best, 5 the worst.
    rank_likelihood(c(1,2,3,5)) + # person 4 says 1 is the best, 5 the worst.
    rank_likelihood(c(2,1,3,4))   # person 5 says 2 is the best, 4 the worst.
H
```

@@

```{r}
L1 <- loglik(indep(equalp(H)),H)  # support at equal strengths
L2 <- loglik(indep(maxp(H)),H)    # maximum support
L2-L1
pchisq(2*(L2-L1),df=1,lower.tail=FALSE) # p-value of asymptotic distribution
```

Thus we reject the hypothesis of equal strength.  Now we can assess
the hypothesis that person `a` does in fact have the highest strength
of the five.  We can follow the reasoning in the _icons_
demonstration:

```{r}
o <- function(Ul,Cl,startp,give=FALSE){
    small <- 1e-4  #  ensure start at an interior point
    if(missing(startp)){startp <- small*(1:4)+rep(0.1,4)}
    out <- maxp(H, startp=small*(1:4)+rep(0.1,4), give=TRUE, fcm=Ul,fcv=Cl)
		if(give){
			return(out)
		} else {
			return(out$value)
		}
}

p2max <- o(c(-1, 1, 0, 0), 0)
p3max <- o(c(-1, 0, 1, 0), 0)
p4max <- o(c(-1, 0, 0, 1), 0)
p5max <- o(c(-2,-1,-1,-1),-1)
```

(the final line is different because `p5` is the fillup value).

```{r}
(likes <- c(p2max,p3max,p4max,p5max))
max(likes)
```

Observe that the hypothesis with the maximum likelihood is
$p_2\geqslant p_1$, as might be expected on the grounds that in the
unconstrained case we have
$\hat{p_2}>\max\left(\hat{p_3},\hat{p_4},\hat{p_5}\right)$.  Now
compare the constrained maximum with the unconstrained maximum:

```{r}
maxp(H,give=TRUE)$value
```

So the support loss by the constraint is

```{r}
(L <- maxp(H,give=TRUE)$value - max(likes))
```

short of Edwards's two-units-of-support criterion.  There is no strong
evidence to support the assertion that person 1 is actually stronger
than person 2 in the sense that $p_1>p_2$.

@@

Suppose that we subsequently observe order statistic `5,4,3,2,1`, that
is, person 5 is the best, 4 is the second best, and so on to 1 being
the worst.  Is this observation consistent with the previous dataset?

@@

```{r}
library(partitions)
imH <- indep(maxp(H))
f <- function(o){loglik(imH,rank_likelihood(o),log=FALSE)}
LL <- apply(perms(5),2,f)
length(LL)
sum(LL)
head(LL)
```

@@

```{r,beforetraining}
H <- hyper2(pnames=c(paste("p",1:5,sep="")))  # 'S' the strength of training
# observations before training:
Hbefore <- H +
    rank_likelihood(c(5,2,3,4,1)) + #  5 is the best, 4 the worst.
    rank_likelihood(c(5,3,2,4,1)) + #  5 is the best, 1 the worst.
    rank_likelihood(c(4,5,3,2,1)) + #  4 is the best, 1 the worst.
    rank_likelihood(c(5,3,4,1,2)) + #  5 is the best, 2 the worst.
    rank_likelihood(c(2,5,3,1,4))   #  2 is the best, 4 the worst.
Hbefore
```

Now, after the training we have two order observations: `1,2,3,4,5`
and `3,1,5,2,4`.  Thus in the first observation, player 1 comes first
and in the second observation, player 1 comes second second (after
player 3).  The likelihood for the first observation would be

@@

```{r,usepwafortraining}
Hafter <- rank_likelihood(1:5)+rank_likelihood(c(3,1,5,2,4))
pnames(Hafter) <- pnames(Hbefore)
Hafter %<>%  pwa("p1")
Hafter
```

>>>>

```{r loadlib}
library(hyper2)

```

@@

```{r data}

H <- hyper2() + 
  race(c("p5","p2","p3","p4")) + # person 1 says p5 is the best, p4 the worst.
  race(c("p1","p5","p3","p4")) + # person 2 says p1 is the best, p4 the worst.
  race(c("p1","p4","p2","p5")) + # person 3 says p1 is the best, p5 the worst.
  race(c("p1","p2","p3","p5")) + # person 4 says p1 is the best, p5 the worst.
  race(c("p2","p1","p3","p4"))   # person 5 says p2 is the best, p4 the worst.
H
```

@@

```{r}
equalp.test(H)
```

Thus we reject the hypothesis of equal strength.  Now we can assess
the hypothesis that person `a` does in fact have the highest strength
of the five.  We can follow the reasoning in the _icons_
demonstration and test $H_0\colon p_1=p_2$:

```{r}
samep.test(H,c("p1","p2"))
```

There is no strong evidence to support the assertion that person 1 is
actually stronger than person 2 in the sense that $p_1>p_2$.

@@

Suppose that we subsequently observe order statistic `p5,p4,p3,p2,p1`,
that is, person 5 is the best, 4 is the second best, and so on to 1
being the worst.  Is this observation consistent with the previous
dataset?

@@

```{r}
library(partitions)
M <- perms(5)
M[] <- paste("p",M,sep="")
imH <- indep(maxp(H))
f <- function(o){loglik(imH,race(o),log=FALSE)}
LL <- apply(perms(5),2,f)
length(LL)
sum(LL)
head(LL)
```

@@

```{r,beforetraining}
H <- hyper2(pnames=c(paste("p",1:5,sep="")))  # 'S' the strength of training
# observations before training:
Hbefore <- H +
  race(c("p5","p2","p3","p4","p1")) + #  5 is the best, 4 the worst.
  race(c("p5","p3","p2","p4","p1")) + #  5 is the best, 1 the worst.
  race(c("p4","p5","p3","p2","p1")) + #  4 is the best, 1 the worst.
  race(c("p5","p3","p4","p1","p2")) + #  5 is the best, 2 the worst.
  race(c("p2","p5","p3","p1","p4"))   #  2 is the best, 4 the worst.
Hbefore
```

Now, after the training we have two finishing order observations:
`p1,p2,p3,p4,p5` and `p3,p1,p5,p2,p4`.  Thus in the first observation,
`p1` comes first and in the second observation, `p1` comes second
second (after `p3`).  The likelihood for the first observation would
be

@@

```{r,usepwafortraining}
Hafter <- race(c("p1","p2","p3","p4","p5")) + race(c("p3","p1","p5","p2","p4"))
Hafter %<>%  pwa("p1")
Hafter
```
